---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# collateral <img src="man/figures/logo.svg" align="right" width="180px" />

<!-- badges: start -->
<!-- badges: end -->

Map complex operations `safely` or `quietly` (or both: `peacefully`!), quickly see the captured side effects, and quickly spot and isolate captured side effects.

The `collateral` package extends the power of [`purrr`'s side effect-capturing functions](https://purrr.tidyverse.org/reference/safely.html), giving you:

* drop-in `map()` variants, allowing you to capture side effects from functions mapped over lists, vectors and list-columns;
* fancy tibble output, allowing you to see which rows delivered errors or side effects; and
* helpers for summarising side effects or filtering tibbles and lists for present side effects.

If you're not familiar with `purrr` or haven't used a list-column workflow in R before, the [`collateral` vignette](https://rensa.co/collateral/articles/collateral.html) shows you how it works, the benefits for your analysis and how `collateral` simplifies the process of handling complex mapped operations.

If you're already familiar with `purrr`, the [tl;dr](https://en.wikipedia.org/wiki/Wikipedia:Too_long;_didn%27t_read) is that [`collateral::map_safely()` and `collateral::map_quietly()` (and their `map2` and `pmap` variants)](https://rensa.co/collateral/reference/collateral_mappers.html) will automatically wrap your supplied function in `safely()` or `quietly()` and will provide enhanced `print()`ed output and tibble displays. You can then use the [`has_*()`](https://rensa.co/collateral/reference/has.html) and [`tally_*()`](https://rensa.co/collateral/reference/tally.html) functions to filter or summarise the returned tibbles or lists.


## Installation

You can install the released version of collateral from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("collateral")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("rensa/collateral")
```
## Example

This example uses the famous `mtcars` dataset---but first, we're going to sabotage a few of the rows by making them negative. The `log` function produces `NaN` with a warning when you give it a neagtive number.

It'd be easy to miss this in a non-interactive script if you didn't explicitly test for the presence of `NaN`! Thankfully, with collateral, you can see which operations threw errors, which threw warnings, and which produced output: 

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(collateral)

test =
  # tidy up and trim down for the example
  mtcars %>%
  rownames_to_column(var = "car") %>%
  as_tibble() %>%
  select(car, cyl, disp, wt) %>%
  # spike some rows in cyl == 4 to make them fail
  mutate(wt = dplyr::case_when(
    wt < 2 ~ -wt,
    TRUE ~ wt)) %>%
  # nest and do some operations peacefully
  nest(data = -cyl) %>%
  mutate(qlog = map_peacefully(data, ~ log(.$wt)))

test
```

Here, we can see that all operations produced output (because `NaN` is still output)---but a few of them also produced warnings! You can then see those warnings...

```{r}
test %>% mutate(qlog_warning = map_chr(qlog, 'warnings', .null = NA))
```

... filter on them...

```{r}
test %>% filter(!has_warnings(qlog))
```

... or summarise them, for either interactive or non-interactive purposes:

```{r}
summary(test$qlog)
```

## Other features

The collateral package is now fully integrated with the `furrr` package, so you can safely and quietly iterate operations across CPUs cores or remote nodes. All collateral mappers have `future_*`-prefixed variants for this purpose.

## Support

If you have a problem with `collateral`, please don't hesitate to [file an issue](https://github.com/rensa/collateral/issues/new) or [contact me](twitter.com/rensa_co)!
